\Title

### 2013-01-23 synmut.tex

HIV evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via killer T cells (CTLs) and neutralizing antibodies
(ABs)~\citep{pantaleo_immunopathogenesis_1996} and facilitated by the high
mutation rate of HIV~\citep{mansky_lower_1995}. When the host develops a CTL or
AB response against a particular HIV epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving part of the HIV genome are the so called
variable loops of the envelope protein gp120, which need to avoid recognition by
neutralizing ABs.  Mutations in \env~, the gene encoding for gp120, spread
through the population within a few months (see \figurename~\ref{fig:aft}, solid
lines).  The (Malthusian) effect size of these beneficial mutations is of the
order of $s_a \sim 0.01$~\citep{neher_recombination_2010}.

Escape mutations are strongly selected for their effect on the amino acid
sequence of the HIV proteins. Conversely, synonymous mutations are commonly
used as approximate neutral markers in studies of viral evolution. Neutral
markers are very useful in practice, because they can be used to make inferences
about the stochastic forces driving evolution~\citep{yang_statistical_2000}.
The viral genome, however, needs to satisfy further constraints in addition to
immune escape, such as efficient processing and translation, nuclear export, and
packaging into the viral capsid: all these processes operate at the RNA level
and are sensitive to synonymous changes. A few functionally important RNA
elements are well characterized. For example, a certain RNA sequence in the HIV
genome, called \rev{} response element (RRE), enhances nuclear export of viral
transcripts~\citep{fernandes_hiv-1_2012}. Another well studied case is the
interaction between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and bound by a specifical pseudoknotted RNA structure in the viral 5'
untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
Nucleotide-level fitness effects have been observed beyond RNA structure as
well. Recent studies have shown that genetically engineered HIV strains with
a skewed codon usage bias towards more or less abundant tRNAs
replicate better or worse, respectively~\citep{ngumbela_quantitative_2008,
li_codon-usage-based_2012}. A similar conclusion has been reached about
influenza, and codon-pessimized influenza strains have been shown to be good
live attenuated vaccines in mice~\citep{mueller_live_2010}. Purifying selection
beyond the protein sequence is therefore expected, but it seems reasonable
that the bulk of positive selection through the immune system be restricted to
amino acid sequences.

@@

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that a substantial fraction of these mutations is deleterious. We
further show that, although such synonymous mutations cannot be used as neutral
markers, the degree to which they hitchhike with nearby nonsynonymous mutations
is very informative; their ability to hitchhike for extended times itself is
rooted in the small recombination rate of
HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
analysis of fixation probabilities to the nonsynonymous mutations, we show that
time dependent selection or strong competition of escape mutations inside the
same epitope are necessary to explain the observed patterns of fixation and
loss.

@@

The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e. with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{fixp},
this is a simple consequence of the fact that
(a) exactly one of the $N$ individuals in the current population will be
the common ancestor of the entire future population at a particular locus and
(b) this ancestor has a probability $\nu$ of carrying the mutation.
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. Time series sequence data enable a
direct observation of both the current frequency $\nu$ of any particular
mutation and its future fate (fixation or extinction). They therefore represent
a simple way to investigate average properties of different classes of
mutations.

\subsection{Synonymous polymorphisms in \env, C2-V5 are mostly deleterious}

@@

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA} in a cumulative histogram, the reactivities of fixed
alleles (red line) are systematically larger than of alleles that are doomed to
extinction (blue line) (Kolmogorov-Smirnov test, $P\approx 0.002$).
In other words, alleles that are likely to be
breaking RNA helices are also more likely to revert and finally be lost from the
population. As a control, the average over non-observed but potentially
available polymorphisms lies between the two curves (green line), as expected
(because only some of them will be helix breakers). Furthermore, as a
complementary analysis, we split the synonymous mutations in the extended V1-V5
region further into conserved and variable regions and found that the biggest
depression in fixation probability is observed in the conserved stems, while the
variable loops show little deviation from the neutral signature, see
\FIG{SHAPEB}.

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness defect of synonymous mutations, in particular codon
usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
human housekeeping genes~\citep{jenkins_extent_2003}. Moreover, codon-optimized
and -pessimized viruses have recently been generated and shown to replicate
better or worse than wild type strains,
respectively~\citep{li_codon-usage-based_2012, ngumbela_quantitative_2008,
coleman_virus_2008}. We do not find, however, any evidence for a contribution of
CUB to the ultimate fate of synonymous alleles. Several lines of thought support
this result. First of all, within a single patient, we do not observe any bias
towards more human-like CUB in the synonymous mutations that reach fixation
rather than extinction. Second, although codon-optimized HIV seems to perform
better {\it in vitro}, the distance in CUB between HIV and human genes is not
shrinking at the macroevolutionary level. Third, it is a common phenomenon for
retroviruses to use variously different codons from their hosts, and CUB effects
on fitness are thought to be so small that divergent nucleotide composition has
been suggested as a possible mechanism for viral
speciation~\citep{bronson_nucleotide_1994}. Fourth, CUB in the V1-C5 region is
not very different from other parts of the HIV genome, whereas the reduced
fixation probability is only observed there. In conclusion, although we cannot
exclude an effect of CUB on fitness as a general rule, we expect it to be a
minor effect in our context.

@@

As far as population genetics models are concerned, our study uncovers the
subtle balance of evolutionary forces governing intrapatient HIV evolution. The
fixation and extinction times and probabilities represent a rich and simple
summary statistics to test sequencing data and computer simulation upon. A
similar method has been recently used in a longitudinal study of
influenza~\citep{strelkowa_clonal_2012}. The propagators suggested in that
paper, however, represent ratios between (certain kinds of) nonsynonymous
mutations and synonymous ones, hence they are inadequate to investigate
synonymous changes themselves. Those authors also conclude that several
beneficial mutations segregate simultaneously in influenza, a scenario
remarkably similar to our within-epitope competition picture. These results
jointly suggest that viral evolution proceeds by multiple concurrent sweeps
rather then by successive fixation~\citep{desai_beneficial_2007, neher_rate_2010}.

Finally, our results emphasize the inadequacy of independent site
models of HIV evolution, especially in the light of transient effects on
sweeping sites, such as time-dependent selection and within-epitope negative
epistasis. With specific regard to these two scenarios, although a final word
about which mechanism is more widespread is yet to be spoken, both intuition
and biological evidence from the literature support a mixed scenario~\citep{richman_rapid_2003,
moore_limited_2009, bar_early_2012}. Note also that, unlike influenza, HIV does
recombine if rarely, hence clonal interference as studied in
ref.~\citep{strelkowa_clonal_2012} is only a short-term effect.

\section{Methods}
\subsection{Sequence data collection}
Longitudinal intrapatient viral RNA sequences were collected for published
studies~\citep{shankarappa_consistent_1999,
liu_selection_2006, bunnik_autologous_2008} and downloaded from the Los Alamos
National Laboratory (LANL) HIV sequence database~\citep{LANL2012}. The sequences from
some patients showed signs of HIV compartimentalization into subpopulations and
were discarded; a grand total of 11
patients with approximately 6 time points each and 10 sequences per time point
were analyzed. The time interval or resolution between two ocnsecutive sequences
was approximately 6 to 18 months.

\subsection{Sequence analisys}
The good sequences were aligned within each patient
via the translated amino acid sequence, using
Muscle~\citep{edgar_muscle:_2004}, and to the NL4-3 reference sequence probed
by \citet{watts_architecture_2009}. Within each patient, a consensus RNA
sequence at the first time point was used to classify alleles as ancestral or
derived at all sites. Problematic sites that included large frequencies of gaps
were excluded from the analysis, because variable regions are known to be
subject to frequent indels, while our analysis is limited to nucleotide
substitutions. Time series of allele frequencies were extracted from the
sequences.

@@

For the correlation analysis with RNA secondary structure, the SHAPE scores were
downloaded from the journal website~\citep{watts_architecture_2009}. By virtue
of the alignment of the longitudinal sequences with the reference used by
\citet{watts_architecture_2009}, SHAPE reactivities were assigned to most sites.
Problematic assignments in indel-rich regions were excluded from the analysis.
In order to restrict the analysis to synonymous polymorphisms, a lower frequency
threshold of 0.15 was used (other thresholds yielded the same results). Since
very few polymorphisms hitchhike beyond, say, a frequency of 0.5, this pool is
enriched for to-be-lost mutations; hence the "lost" curve in \FIG{SHAPEA}
contains much more points than the "fixed" one.

>>>>

HIV evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via killer T cells (CTLs) and neutralizing antibodies
(ABs)~\citep{rambaut_causes_2004} and facilitated by the high
mutation rate of HIV~\citep{mansky_lower_1995}. When the host develops a CTL or
AB response against a particular HIV epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving part of the HIV genome are the so called
variable loops of the envelope protein gp120, which need to avoid recognition by
neutralizing ABs.  Mutations in \env, the gene encoding for gp120, spread
through the population within a few months (see \figurename~\ref{fig:aft}, solid
lines).

These escape mutations are strongly selected for their effect on the amino acid
sequence of the viral proteins. Conversely, synonymous mutations are commonly
used as approximately neutral markers in studies of viral evolution. Neutral
markers are very useful in practice, because there evolution can be compared to
that of putatively functional sites to detect purifying or directional selection
\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In most species,
however, it is  known that synonymous codons don't evolve completely neutrally
but that some codons are favored over others \citep{plotkin_synonymous_2011}.
In addition to maintaining protein function and avoiding immune
recognition, the HIV genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate at the RNA
level and are sensitive to synonymous changes. A few functionally important RNA
elements are well characterized. For example, a certain RNA sequence in the HIV
genome, called \rev{} response element (RRE), enhances nuclear export of viral
transcripts~\citep{fernandes_hiv-1_2012}. Another well studied case is the
interaction between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and bound by a specifical pseudoknotted RNA structure in the viral 5'
untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
Nucleotide-level fitness effects have been observed beyond RNA structure as
well. Recent studies have shown that genetically engineered HIV strains with
a skewed codon usage bias towards more or less abundant tRNAs
replicate better or worse, respectively~\citep{ngumbela_quantitative_2008,
li_codon-usage-based_2012}. A similar conclusion has been reached about
influenza, and codon-pessimized influenza strains have been shown to be good
live attenuated vaccines in mice~\citep{mueller_live_2010}. Purifying selection
beyond the protein sequence is therefore expected, but it seems reasonable
that the bulk of positive selection through the immune system be restricted to
amino acid sequences.

@@

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that a substantial fraction of these mutations is deleterious. We
further show that, although such synonymous mutations cannot be used as neutral
markers, the degree to which they hitchhike with nearby nonsynonymous mutations
is very informative both about their on deleterious effect, as well as the
strength and nature of selection for nearby positively selected sites.
Their ability to hitchhike for extended times, which is a core requirement for
our analysis, is rooted in the small recombination rate of
HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
analysis of fixation probabilities to the nonsynonymous mutations, we show that
time dependent selection or strong competition of escape mutations inside the
same epitope are necessary to explain the observed patterns of fixation and
loss.

@@

The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e. with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{fixp},
this is a simple consequence of the fact that
(a) exactly one of the $N$ individuals in the current population will be
the common ancestor of the entire future population at a particular locus and
(b) this ancestor has a probability $\nu$ of carrying the mutation.
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. Time series sequence data enable a
direct observation of both the current frequency $\nu$ of any particular
mutation and its future fate (fixation or extinction).

\subsection{Synonymous polymorphisms in \env, C2-V5, are mostly deleterious}

@@

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA} in a cumulative histogram, the reactivities of fixed
alleles (red line) are systematically larger than of alleles that are doomed to
extinction (blue line) (Kolmogorov-Smirnov test, $P\approx 0.002$).
In other words, alleles that are likely to be
breaking RNA helices are also more likely to revert and finally be lost from the
population. As a control, the average over non-observed but potentially
available polymorphisms lies between the two curves (green line), as expected
(because only some of these mutations will interfere with stem loop formation).
To test the hypothesis that mutations in C2-V5 are lost since break stems
in the conserved between the variable loops, we split the synonymous mutations
in the extended V1-V5 region further into conserved and variable regions and
found that the biggest depression in fixation probability is observed in the
conserved stems, while the variable loops show little deviation from the
neutral signature, see \FIG{SHAPEB}. This is consistent with important stem
structures in conservered regions between loops.

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness defect of synonymous mutations, in particular codon
usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
human codons~\citep{jenkins_extent_2003}. Moreover, codon-optimized
and -pessimized viruses have recently been generated and shown to replicate
better or worse than wild type strains,
respectively~\citep{li_codon-usage-based_2012, ngumbela_quantitative_2008,
coleman_virus_2008}. We do not find, however, any evidence for a contribution of
CUB to the ultimate fate of synonymous alleles. Several lines of thought support
this result. First of all, within a single patient, we do not observe any bias
towards more human-like CUB in the synonymous mutations that reach fixation
rather than extinction. Second, although codon-optimized HIV seems to perform
better {\it in vitro}, the distance in CUB between HIV and human genes is not
shrinking at the macroevolutionary level. Third, it is a common phenomenon for
retroviruses to use variously different codons from their hosts, and CUB effects
on fitness are thought to be so small that divergent nucleotide composition has
been suggested as a possible mechanism for viral
speciation~\citep{bronson_nucleotide_1994}. Fourth, CUB in the V1-C5 region is
not very different from other parts of the HIV genome, whereas the reduced
fixation probability is only observed there. In conclusion, although we cannot
exclude an effect of CUB on fitness as a general rule, we expect it to be a
minor effect in our context.

@@

Our study uncovers the
subtle balance of evolutionary forces governing intrapatient HIV evolution. The
fixation and extinction times and probabilities represent a rich and simple
summary statistics useful to characterize sequencing data and compare to
models via computer simulations.
A similar method has been recently used in a longitudinal study of
influenza~\citep{strelkowa_clonal_2012}. The propagators suggested in that
paper, however, represent ratios between nonsynonymous
mutations and synonymous ones, hence they are inadequate to investigate
synonymous changes themselves. The authors also conclude that several
beneficial mutations segregate simultaneously in influenza, a scenario
remarkably similar to our within-epitope competition picture. While
recombination hardly affects evolution within an epitope, it might facilitate
immune escape by easing competition between different epitopes
\citep{neher_rate_2010,Rouzine:2005p17398}.

Our results emphasize the inadequacy of independent site
models of HIV evolution and the common assumption that selection is time
independent. If genetic variation is only transiently beneficial, existing
methods to quantify selection will yield substantial underestimates
\citep{williamson_adaptation_2003,neher_rate_2010,OTHER}. To explain the
observations regarding the fixation probabilities of non-synonymous mutations,
either transient selection, or substational within-epitope competition are
necessary. Which mechanism is more widespread is not clear as of now,
there is evidence for both~\citep{richman_rapid_2003, moore_limited_2009,
bar_early_2012}.

\section{Methods}
\subsection{Sequence data collection}
Longitudinal intrapatient viral RNA sequences were collected for published
studies~\citep{shankarappa_consistent_1999,
liu_selection_2006, bunnik_autologous_2008} and downloaded from the Los Alamos
National Laboratory (LANL) HIV sequence database~\citep{LANL2012}. The sequences from
some patients showed signs of HIV compartimentalization into subpopulations and
were discarded; a grand total of 11
patients with approximately 6 time points each and 10 sequences per time point
were analyzed. The time interval or resolution between two consecutive sequences
was approximately 6 to 18 months.

\comment{How did you determine who to discard. Maybe we should include the PCA
plots into the supplement.}

\subsection{Sequence analysis}
The good sequences were aligned within each patient
via the translated amino acid sequence, using
Muscle~\citep{edgar_muscle:_2004}, and to the NL4-3 reference sequence used
by \citet{watts_architecture_2009} in the SHAPE assay. Within each patient, a
consensus RNA sequence at the first time point was used to classify alleles as ancestral or
derived at all sites. Problematic sites that included large frequencies of gaps
were excluded from the analysis to avoid artefactual substitutions due to
alignment errors. Time series of allele frequencies were extracted from the
sequences.

@@

For the correlation analysis with RNA secondary structure, the SHAPE scores were
downloaded from the journal website~\citep{watts_architecture_2009}. By virtue
of the alignment of the longitudinal sequences with the reference used by
\citet{watts_architecture_2009}, SHAPE reactivities were assigned to most sites.
Problematic assignments in indel-rich regions were excluded from the analysis.
In order to restrict the analysis to synonymous polymorphisms, a lower frequency
threshold of 0.15 was used (other thresholds yielded the same results). Since
very few polymorphisms hitchhike beyond, say, a frequency of 0.5, this pool is
enriched for to-be-lost mutations; hence the ``lost" curve in \FIG{SHAPEA}
contains much more points than the ``fixed" one.
